Viscous Transport in Eroding Porous Media

### 2020-01-08 docs/transport/Response_to_Reviewer_1.tex

  \item After equation (5.18), we now state:
    \begin{quotation}
      \noindent
      These quadrature formula are also not in a Barycentric form
      because of catastrophic cancellation in the numerator. As is the
      case when computing $v′(x)$, this error is sufficiently small at
      the geometries and resolutions we consider, so we use these
      quadrature formulas.
    \end{quotation}

  \item To make it clear which quadrature methods we apply, we have
    added the following paragraph to the end of Section 5.2:
    \begin{quotation}
      \noindent
      In section 6, the quadrature rules (5.6), (5.7), (5.9), (5.10),
      (5.15), and (5.18) are used to form simulations of nearly-touching
      eroding grains, and to study dynamics of the flow in regions
      arbitrarily close to eroding grains.
    \end{quotation}

  \item In the second paragraph of Section 5.2, we updated the last
    sentence to read:
    \begin{quotation}
      \noindent
      After briefly summarizing this method, we compute the second
      derivative so that the shear stress and vorticity can be
      accurately computed.
    \end{quotation}

  \item We start Section 5.3 with the following revised sentence:
    \begin{quotation}
      \noindent
      By using these quadrature rules, the velocity, shear stress, and
      vorticity are calculated at the target points to simulate erosion
      and transport.
    \end{quotation}

@@

\end{itemize}

\noindent
\comment{In my opinion, the claim of having developed a ``new quadrature
method" is too strong. After all, there is nothing new about the
barycentric quadrature itself. Rather, I suggest that maybe a more
appropriate claim would be that the authors have developed a ``new
numerical scheme" for the Stokes porous media flow, where the accurate
quadrature is one ingredient in this scheme.}
\begin{itemize}
  \item We have toned down some of the claims. The following modified
    paragraphs and sentences are in the revised manuscript:
    \begin{quotation}
      \noindent
      In previous work (Quaife \& Moore 2018), the BIE is discretized
      with the trapezoid rule, and this is acceptable if the grains are
      sufficiently separated (Trefethen \& Weideman 2014). However, when
      grains are close, the result is a nearly-singular integrand and
      the error of the trapezoid rule is no longer acceptable. One of
      the earliest quadrature methods for nearly-singular integrands was
      developed by Baker \& Shelley (1986), and in recent years, many
      other schemes have followed (af Klinteberg \& Tornberg 2018;
      Helsing \& Ojala 2008; Beale {\em et al.}~2016; Beale \& Lai 2001;
      Kl\"{o}ckner {\em et al.}~2013). We use a Barycentric quadrature
      method (Barnett 2014; Barnett {\em et al.}~2015) since it is a
      non-intrusive modification of the trapezoid rule, and the error is
      guaranteed to be uniformly bounded. We use this quadrature method
      to compute the velocity gradient, which is needed to evaluate the
      shear stress and the fluid vorticity.
    \end{quotation}

@@

\noindent
\comment{The porosity-tortuosity models (4.6c \& 4.6d) are not
explicitly compared in the results section. How much worse are they
comparing to the other models?}
\begin{itemize}
  \item We have adjusted each of the appropriate sections.

  \item In Section 6.2, we have added the following sentence:
    \begin{quotation}
      \noindent
      A slightly larger root-mean-square error of $6.70 \times 10^{-3}$
      is possible with the logarithmic porosity-tortuosity model (4.6b),
      but the other two models have a root- mean-square error that is an
      order of magnitude larger.
    \end{quotation}

  \item In Section 6.3, we have added the following sentence:
    \begin{quotation}
      \noindent
      However, the root-mean-square error of the other three models is
      bounded by $2.38 \times 10^{-2}$.
    \end{quotation}

  \item In Section 6.4, we have added the following sentence:
    \begin{quotation}
      \noindent
      We also compute the lines of best fit using the
      porosity-tortuosity models (4.6).  Each of the models result in a
      root-mean-squared error ranging from $4.60 \times 10^{-3}$ for
      model (4.6d) to $7.20 \times 10^{-3}$ for model (4.6c). The black
      dashed line in figure 15(a) is the line of best fit
      $\widehat{T}(\phi) = \phi^{-0.2459}$, with a root-mean-square
      error of $5.50 \times 10^{-3}$.  We note a slightly better
      root-mean-square error of $4.60 \times 10^{-3}$ is possible with
      the model $\widehat{T}(\phi) = \left(1 + 0.1499(1-\phi)\right)^2$.
    \end{quotation}

@@

\noindent
\comment{When first introduced in Sec 4.3, the ``pore sizes" are not
clearly defined. (My first impression was that pore size = portion of an
edge of the Delaunay diagram. Only at the end in Sec 6.4 do we know that
it is defined as the closest distance between grains.)}
\begin{itemize}
  \item In Figure 3, we clarify the definition by adding the statement:
    {\em represented by the length of the red line segments}.
  \item In the text, we now state:
    \begin{quotation}
    \noindent
      Finally, we define the pore sizes to be the minimum distance
      between all pairs of neighboring grains, and these distances are
      represented by the red line segments in figure 3(a).
    \end{quotation}
\end{itemize}

\noindent
\comment{``Porosity" is undefined throughout the paper.}
\begin{itemize}
  \item We have now formally defined the porosity by rewriting the
    description of coarse-grained models as:
  \begin{quotation}
  \noindent
    Examples of coarse-grained models for porous media flow include
    tortuosities (Matyka {\em et al.}~2008; Duda {\em et al.}~2011;
    Koponen {\em et al.}~1996), geometry connectivity (Knudby \& Carrera
    2005), anomalous dispersion (Dentz {\em et al.}~2004), and
    permeability-porosity relationships (Dardis \& McCloskey 1998;
    Carman 1937), where the porosity is the fraction of the geometry
    that is fluidized.
  \end{quotation}

\end{itemize}

\noindent
\comment{``Area fraction" is plotted in (and only in) Figure 7, while
the rest of the paper used porosity. Maybe plotting porosity here would
give a more consistent and smoother reading experience?}
\begin{itemize}
  \item We examined the area fraction and flow rate in figure 7 so that
    it could be compared to results from our previous work. The goal is
    to demonstrate that we now are able to resolve smaller area
    fractions and correspondingly lower flow rates.

### 2020-01-08 docs/transport/Response_to_Reviewer_2.tex

\section*{Response to Reviewer 2}

\noindent
\comment{equations 2.1.  I am confused as to how the no-slip condition
is implemented as zero velocity when in fact the interface is receding
at a rate Vn.}
\begin{itemize}
  \item This comes from a quasi-steady approximation that we have used
    in our previous work as well as by others. For example, see Section
    5.2 from Mitchell and Spagnolie's JCP paper {\em A generalized
    traction integral equation for Stokes flow, with applications to
    near-wall particle mobility and viscous erosion}.

  \item In Section 2, we now state:
    \begin{quotation}
      \noindent
      Since the rate of erosion is much slower than the background flow,
      a quasi-steady approximation justifies the no-slip boundary
      condition.
    \end{quotation}
\end{itemize}

\noindent
\comment{end of section 4.1.  It seems that this protocol (matching
velocity but not position for reinsertion) results in each particle
moving through a different landscape.  Can the authors comment on the
significance of this?}
\begin{itemize}
  \item The reinsertion approach we take has been shown by others to not
    introduce any significantly artificial decorrelation.

@@

\noindent
\comment{p.17 ``single body close to a wall", the 'experiment' mentioned
here is, I believe, a numerical `experiment'.  It might be good to
indicate that clearly.}
\begin{itemize}
  \item The reviewer is correct that this is a numerical experiment.

  \item There were other instances in the paper where we claimed to be
    performing an `experiment'. We have clarified each of the instances
    by referring to them as a `numerical experiment'.
\end{itemize}

\noindent
\comment{sections 6.2 and 6.3.  How were these geometries of 20 grains
chosen?  Both the sizes and positions of these grains vary between
examples - it's not just the porosity that is different.}
\begin{itemize}
  \item The initial placement and radii of the bodies are chosen with a
    randomized method. The algorithm is descried at the start of Section
    6.

  \item We have updated the start of section 6.2 to read:
    \begin{quotation}
      \noindent
      We randomly initialize 20 eroding grains with an initial porosity
      of $\phi = 37.68\%$ and the Hagen-Poiseuille flow
      $\mathbf{U}(\mathbf{x}) = U (1-y^2, 0)$ imposed on $\Gamma$.
    \end{quotation}

@@

  \item We have updated the start of section 6.4 to read:
    \begin{quotation}
      \noindent
        As a final example, we consider 100 randomly initialized eroding
        bodies with an initial porosity near $50\%$.
    \end{quotation}

\end{itemize}

\noindent
\comment{Figure 8.  Due to what I presume is an artifact of plotting, it
looks like lots of streamlines suddenly end, which is confusing.}
\begin{itemize}
  \item In figure 8, we freeze an eroded geometry and plot the
    trajectory of passive tracers at five different time steps. Since
    the flow is steady and the tracers are passive, the trajectories
    trace out streamlines.

  \item We have updated the caption to read:
    \begin{quotation}
      \noindent
      200 particle trajectories in the second geometry from figure 6.
      The particles are initially equispaced at $(−1,y)$, where $y \in
      (-1, 1)$. The first five snapshots are equispaced in time. The
      bottom right frame is a magnification of the fifth snapshot, but
      with additional trajectories.
    \end{quotation}

@@

  \item We have also updated the paragraph describing figure 8 to
    clarify that it is showing tracer trajectories that trace out
    streamlines.
\end{itemize}

\noindent
\comment{Figure 9.  While the sharp features in subfigure (b) are
apparently finite cusps, they raise the question of whether any infinite
dwell times (divergent $\tau$) are ever observed in these simulations -
i.e.~are particles ever trapped inside recirculating zones?}
\begin{itemize}
  \item Thank you for pointing out this possibility. The
    recirculation zones are close to the eroding bodies and are
    typically very small. Because we measure the local tortuosity by
    initiating streamlines significantly far from the grains, each of
    them connect the inlet to the outlet.

  \item We have updated the text to read:
    \begin{quotation}
      \noindent
      We note that if a streamline originates in a recirculation zone,
      which is possible in two dimensions (Higdon 1985), then the local
      tortuosity will be infinite. However, in each of our numerical
      experiments, all the streamlines we consider connect the inlet to
      the outlet.  Having only finite local tortuosities, the hydraulic
      tortuosity is defined by taking the average over all points on the
      inlet cross-section.
    \end{quotation}
\end{itemize}

\noindent
\comment{p.26 bottom paragraph typo: ``an ... media"}
\begin{itemize}
  \item This typo has been fixed.
\end{itemize}

### 2020-01-08 docs/transport/Response_to_Reviewer_3.tex

\begin{document}

\noindent
Dear Silas Alben,
\\ \\
\noindent
Thank you for handling the manuscript, and we would like to thank the
reviewer for their positive and constructive comments.  The attached
manuscript addresses all the reviewer's comments and the result is what
we believe is a stronger manuscript.  We also note that formatting
issues were resolved in the bibliography. An itemized list of the
changes addressing the reviewer's comments are below.
\\ \\
\noindent
Sincerely,
\\
\noindent
Shang-Huan Chiu, Nick Moore, and Bryan Quaife

\section*{Response to Reviewer 3}

\noindent
\comment{Pores in natural porous media care of three-dimensional
geometry.  The authors may include some discussion on how to extend
their numerical method to three-dimensional problems.  In addition, they
may also discuss the limitations of the erosion analysis in
2-dimensional flow to be applied for three-dimensional flow. If the
erosion process in three-dimensional flow is very different from that in
two-dimensional flow, the analysis of this paper may have less relevant
to the analysis of erosion in natural porous media.}
\begin{itemize}
  \item The reviewer is correct that by assuming the geometry is
    two-dimensional, we have limited the applicability for natural
    porous media. There are several numerical challenges to simulate
    three-dimensional erosion, with the most significant being the
    $\theta$--$L$ formulation, quadrature method, spectral
    representation of the geometry, and the boundary integral equation
    solver.

  \item In three-dimensions, others have extended the $\theta$--$L$
    formulation to remove stiffness in interfacial flow, spectrally
    represented geometries, and efficiently solving BIEs.  The challenge
    with these methods is their computational cost. To consider
    suspensions with the resolution and number of grains that we
    consider, the BIE would most likely require a parallel
    implementation that would run on high-performance computing
    architectures. At this point, this is not the goal of our work.
    Rather, we are interested in using very high-order methods to
    provide an analysis of the effects of erosion on transport which
    seems to be missing from the literature.

@@

  \item The Barycentric quadrature method relies on a complex variable
    formulation, so it is inherently two-dimensional. However, there are
    other quadrature methods that can be applied in three dimensions,
    but they are also computationally expensive. For example, if the
    geometry is represented with spherical harmonics, a
    Gaussian-Legendre quadrature method requires $\mathcal{O}(p^3)$
    operations, where $p$ is the highest-order spherical harmonic.

  \item The manuscript has been modified as discussed in the response to
    the reviewer's second comment.
\end{itemize}

@@

\noindent
\comment{In the numerical examples of the paper the grains are all
separated. In natural porous media, grains are compressed to form grain
contacts. The authors may include discussion on numerical challenges in
studying erosion process in natural porous media.}
\begin{itemize}
  \item The reviewer is correct that porous media grains are compressed
    and form grain contact. This is the result of processes such as
    sedimentation and transport of the grains themselves. This, again,
    is a limitation of our method since we only consider immobile
    bodies.

  \item We plan to consider mobile bodies in the future, and this will
    require additional computational effort.

  \item To the best of our knowledge, nobody has studied the effect
    erosion has on transport through a porous media. Therefore, as a
    first step, we limit ourselves to two-dimensional suspensions of
    immobile bodies, and we plan to remove these limitations in the
    future. In the Introduction, we have added the following paragraph:
    \begin{quotation}
      \noindent
      Since high-order methods have never been used to study transport
      in eroding porous media, we consider, as a first step,
      two-dimensional and immobile grains. These assumptions allow us to
      simulate large densely-packed suspensions (up to $100$ in this
      manuscript), and to clearly interpret the results. More realistic
      three-dimensional and mobile configurations would require
      overcoming a few challenges. First, a three-dimensional extension
      of the $\theta$--$L$ formulation exists (Ambrose {\em et
      al.}~2013), but the Barycentric quadrature method used here relies
      on a complex variable formulation of the BIE, making it inherently
      two-dimensional. Therefore, a different quadrature method would
      be required. Additional challenges include spectral
      representations of three-dimensional geometries and efficiently
      solving discretizations of the BIE. While high-order methods for
      three-dimensional BIEs have been developed, their computational
      cost limits the number of bodies that can be practically
      simulated. Lastly, mechanisms such as sedimentation play a role in
      realistic media and affect the pore structure. Simulating mobile
      grains introduces the additional challenge of resolving contact,
      which we plan to address in future work.
    \end{quotation}
\end{itemize}

### 2020-01-08 docs/transport/erosion_transport.tex
finished up all but one revision for the transport paper

To compute stable simulations of erosion, we use methods of high-order
in both space and time. The time integration is unchanged from previous
work~\citep{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be
triggered by changes in sign of the shear stress, and we use a stable
second-order Runge-Kutta method applied to the {\thL}
coordinates~\citep{hou-low-she1994} of the eroding grains. In previous
work~\citep{qua-moo2018}, the BIE is discretized with the trapezoid
rule, and this is acceptable if the grains that are sufficiently
separated~\citep{tre-wei2014}. However, when grains are close, the
result is a nearly-singular integrand and the error of the trapezoid
rule is no longer acceptable. One of the earliest quadrature methods for
nearly-singular integrands was developed by~\citet{bak-she1986}, and in
recent years, many other schemes have followed~\citep{kli-tor2018,
hel-oja2008a, bea-yin-wil2016, bea-lai2001, klo-bar-gre-one2013}.  We
use a Barycentric quadrature method~\citep{bar2014, bar-wu-vee2015}
since it is a non-intrusive modification of the trapezoid rule, and the
error is guaranteed to be uniformly bounded.  We use this quadrature
method to compute the velocity gradient, which is needed to evaluate the
shear stress and the fluid vorticity.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity is a dimensionless number that quantifies the amount of
twisting of streamlines. Unlike the dispersion calculations, we do not
use reinsertion to form long trajectories.  The local tortuosity is
\begin{align}
  \tau(y_0) = \frac{\lambda(y_0)}{d}.
  \label{eqn:localTort}
\end{align}
Here the streamline originates on the inlet cross-section $x=x_0$ at
$(x_0,y_0)$, and its arclength, $\lambda(y_0)$, is calculated until the
streamline passes the parallel outlet cross-section $x = x_0 + d$.  In
this work, we consider streamlines originating at $x=-1$ and terminating
at $x=1$, so $d=2$.  However, other choices for the terminal point when
computing the tortuosity are sometime used~\citep{dud-koz-mat2011}.
We note that if a streamline originates in a recirculation zone, which
is possible in two dimensions~\citep{hig1985}, then the local tortuosity
will be infinite. However, in each of our numerical experiments, all the
streamlines connect the inlet to the outlet. Having only finite local
tortuosities, the hydraulic tortuosity is defined by taking the average
over all points on the inlet cross-section
\begin{align}
  T = \frac{1}{d}\left(\int_{S}u_1(x_0,y_0)\lambda(y_0)\,dy_0 \right)
  \Bigg/
  \left(\int_{S}u_1(x_0,y_0)\,dy_0 \right),
  \label{eqn:tortuosity1}
\end{align} 
where $S$ is the inlet cross-section $x = -1$, $u_1(y_0)$ is the
$x$-component of the velocity at the initial point of the streamline,
and $d=2$ is the distance between the inlet and outlet.  Note that $T
\geq 1$, and $T=1$ only if no grains are present.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By using these quadrature rules, the velocity, shear stress, and
vorticity are calculated at the target points necessary to simulate
erosion and transport. However, applied directly, it requires $O(N^2)$
operations, where $N$ is the total number of source and target points.
By using a fast summation method, such as the fast multipole method
(FMM)~\citep{gre-rok1987}, the cost can be reduced to $O(N)$ operations.
However, each application of the Barycentric quadrature rules involve
several $N$-body calculations, rendering the computational cost
prohibitive, so we introduce a hybrid method that combines the
Barycentric quadrature rule and an accelerated trapezoid rule.  Note
that the source points of the layer potential is always one of the
eroding bodies or the outer wall, but the target point can either be on
another component of $\bd\Omega$ or it can be in the fluid bulk
$\Omega$.

>>>>

To compute stable simulations of erosion, we use methods of high-order
in both space and time. The time integration is unchanged from previous
work~\citep{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be
triggered by changes in sign of the shear stress, and we use a stable
second-order Runge-Kutta method applied to the {\thL}
coordinates~\citep{hou-low-she1994} of the eroding grains. In previous
work~\citep{qua-moo2018}, the BIE is discretized with the trapezoid
rule, and this is acceptable if the grains are sufficiently
separated~\citep{tre-wei2014}. However, when grains are close, the
result is a nearly-singular integrand and the error of the trapezoid
rule is no longer acceptable. One of the earliest quadrature methods for
nearly-singular integrands was developed by~\citet{bak-she1986}, and in
recent years, many other schemes have followed~\citep{kli-tor2018,
hel-oja2008a, bea-yin-wil2016, bea-lai2001, klo-bar-gre-one2013}.  We
use a Barycentric quadrature method~\citep{bar2014, bar-wu-vee2015}
since it is a non-intrusive modification of the trapezoid rule, and the
error is guaranteed to be uniformly bounded.  We use this quadrature
method to compute the velocity gradient, which is needed to evaluate the
shear stress and the fluid vorticity.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity is a dimensionless number that quantifies the amount of
twisting of streamlines. Unlike the dispersion calculations, we do not
use reinsertion to form long trajectories.  The local tortuosity is
\begin{align}
  \tau(y_0) = \frac{\lambda(y_0)}{d}.
  \label{eqn:localTort}
\end{align}
Here the streamline originates on the inlet cross-section $x=x_0$ at
$(x_0,y_0)$, and its arclength, $\lambda(y_0)$, is calculated until the
streamline passes the parallel outlet cross-section $x = x_0 + d$.  In
this work, we consider streamlines originating at $x=-1$ and terminating
at $x=1$, so $d=2$.  However, other choices for the terminal point when
computing the tortuosity are sometime used~\citep{dud-koz-mat2011}.  We
note that if a streamline originates in a recirculation zone, which is
possible in two dimensions~\citep{hig1985}, then the local tortuosity
will be infinite. However, in each of our numerical experiments, all the
streamlines we consider connect the inlet to the outlet. Having only
finite local tortuosities, the hydraulic tortuosity is defined by taking
the average over all points on the inlet cross-section
\begin{align}
  T = \frac{1}{d}\left(\int_{S}u_1(x_0,y_0)\lambda(y_0)\,dy_0 \right)
  \Bigg/
  \left(\int_{S}u_1(x_0,y_0)\,dy_0 \right),
  \label{eqn:tortuosity1}
\end{align} 
where $S$ is the inlet cross-section $x = -1$, $u_1(y_0)$ is the
$x$-component of the velocity at the initial point of the streamline,
and $d=2$ is the distance between the inlet and outlet.  Note that $T
\geq 1$, and $T=1$ only if no grains are present.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By using these quadrature rules, the velocity, shear stress, and
vorticity are calculated at the target points to simulate erosion and
transport. However, applied directly, it requires $O(N^2)$ operations,
where $N$ is the total number of source and target points.  By using a
fast summation method, such as the fast multipole method
(FMM)~\citep{gre-rok1987}, the cost can be reduced to $O(N)$ operations.
However, each application of the Barycentric quadrature rules involve
several $N$-body calculations, rendering the computational cost
prohibitive, so we introduce a hybrid method that combines the
Barycentric quadrature rule and an accelerated trapezoid rule.  Note
that the source points of the layer potential is always one of the
eroding bodies or the outer wall, but the target point can either be on
another component of $\bd\Omega$ or it can be in the fluid bulk
$\Omega$.
